---
title: "bird_quant_analysis"
author: "Jackson Barratt Heitmann"
date: "4/4/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

So this first section I'm taking my raw data, and manipulating it into a community
matrix that I can perform some statistics on afterward. 

```{r}
library(lattice)
library(ggplot2)
library(tidyr)
library(dplyr)
library(readr)
library(mobr)
library(vegan)
library(devtools)
install.packages("devtools")
install.packages("devtools.zip", repos = NULL)
library(remotes)
?devtools

#all the bird data together

dat_all <- read.csv("raw_data - all_bird_data.csv")

# env matrix
bird_env <- read.csv("raw_data - veg_2022.csv")

# wetland assessment env matrix

bird_wet_env <- read.csv("raw_data - just_wet_attribute.csv")

# renaming the env dataframe for uniola database

super_bird_matrix <- merge(dat_all, bird_env, by = 'wetland_id')

super_bird_matrix_wet <- merge(super_bird_matrix, bird_wet_env, by = 'wetland_id')


# all data together, need to make super matrix, merge()
# with env data, then separate out, individually 
# env matrix, needs site, year, site_type, etc. 

id3 <- with(super_bird_matrix, paste(wetland_id, date, year))
comm3 <- with(super_bird_matrix_wet, tapply(total, list(id3, species), sum))
comm3 <- ifelse(is.na(comm3), 0, comm3)

# create columns for the group variables by splitting row names up
comm3 <- data.frame(comm3)

#I need to fi this in above code, for some reason it summed 4x ??
comm4 <- comm3/4

# comm3$wetland_id <- sapply(strsplit(row.names(comm3), ' '), function(x) x[1])
# comm3$date <- sapply(strsplit(row.names(comm3), ' '), function(x) x[2])
# comm3$year <- sapply(strsplit(row.names(comm3), ' '), function(x) x[3])

env <- super_bird_matrix_wet %>% select('wetland_id', 'block.y', 'date.x', 'year', 'spp', 'dbh', 'dist', 'site.y', 'site_type.x', 'canopy_cover', 'canopy_dom', 'midstory_cover', 'midstory_dom', 'groundcover', 'dom_herb_group', 'wet_area', 'wet_perimeter')


##making the env matrix properly, with spp_richness as a column, with average dbh and dist

env$site_type.x <- as.factor(env$site_type.x)

nrow(distinct(env))
summary(env$site_type.y)

env_species_counts <- env %>% group_by(wetland_id, site.y, year, date.x, block.y, canopy_cover, canopy_dom, midstory_cover, midstory_dom, groundcover, dom_herb_group, wet_area, wet_perimeter) %>% summarise(
  spp_counts = length(unique(spp)),
  mean_dbh = mean(dbh, na.rm = T),
  mean_dist = mean(dist, na.rm = T),
  site_type = unique(as.character(site_type.x)))


summary(env$site_type.y)

## making the row names the same
rownames(env_species_counts) <- rownames(comm4)

## do some vegan analysis

rda_tree <- rda(comm4 ~ env_species_counts$spp_counts + env_species_counts$mean_dbh + env_species_counts$canopy_cover + env_species_counts$canopy_dom, data=env_species_counts)
rda_tree

RsquareAdj(rda_tree)

plot(rda_tree, type='n', scaling=1)
orditorp(rda_tree, display='sp', cex=0.5, scaling=1, col='blue')
text(rda_tree, display='cn', col='red')


anova(rda_tree, by='margin', permutations=1000)


# making a mob with env and comm matrix

bird_mob <- make_mob_in(comm4, env_species_counts)


mob_test <- get_delta_stats(bird_mob, env_var = 'site_type', ref_level = 'upland',
                            type = 'discrete', tests = c('SAD', 'N'), n_perm = 199)
                                
plot(mob_test)
# making the alpha, beta, gamme plots, these are now obsolete
# they don't work

# bird_stats <- get_mob_stats(bird_mob, group_var = "site_type", ref_level = 'upland', n_perm = 199)


plot(bird_stats, 'S')
plot(bird_stats, 'N')
plot(bird_stats, 'S_n')
plot(bird_stats, 'S_PIE')

# trying to calculate alpha diversity and beta diversity to export
?mobr
bird_div_metrics_S <- calc_comm_div(comm4, 'S', scales = c("alpha"))

bird_div_metrics_S_PIE <- calc_comm_div(comm4, 'S_PIE', scales = c("alpha"))

bird_div_metrics_S_b <- calc_comm_div(comm4, 'S', scales = c("beta"))

bird_div_metrics_S_PIE_b <- calc_comm_div(comm4, 'S_PIE', scales = c("beta"))
 
bird_div_metrics_S_g <- calc_comm_div(comm4, 'S', scales = c("gamma"))

rownames(bird_div_metrics_S) <- rownames(comm4)
rownames(bird_div_metrics_S_PIE) <- rownames(comm4) 
rownames(bird_div_metrics_S_b) <- rownames(comm4)
rownames(bird_div_metrics_S_PIE_b) <- rownames(comm4) 

# try to merge these statistics with my env dataframe, 
# to do some regression modelling

env_matrix <- cbind(env_species_counts, bird_div_metrics_S$value, bird_div_metrics_S_PIE$value)
names(env_matrix)[18] <- 'bird_div_S_a'
names(env_matrix)[19] <- 'bird_div_S_PIE_a'

#checking for normality, bird_div_S_a is good, but S_PIE needs
# a log transformation

hist(env_matrix$bird_div_S_a)

hist(env_matrix_wet$wet_area)

shapiro.test(env_matrix_wet$wet_area)
qqplot(canopy_model$residuals, canopy_model$fitted.values)

# making NAs into 0s
env_matrix[is.na(env_matrix)] <- 0

# need to subset to just get wetlands for some analysis

env_matrix_wet <- subset(env_matrix, site_type == 'wetland')

# sub-setting the JUST wetlands by site, to compare within site variation

env_matrix_halidon <- subset(env_matrix_wet, site.y == 'halidon')

# sub-setting the JUST wetlands by site, to compare within site variation

env_matrix_stono <- subset(env_matrix_wet, site.y == 'stono')


# Model that comparies bird div between site and site type
summary(aov(bird_div_S_a ~ site_type + site.y + canopy_cover, data = env_matrix))


#plot between wetland v upland
ggplot(data = env_matrix, mapping = aes(site_type, bird_div_S_a), color = site_type) + geom_boxplot() + theme_bw()

# plot between stono v halidon
ggplot(data = env_matrix, mapping = aes(site.y, bird_div_S_a), color = site.y) + geom_boxplot() + theme_bw()

# creating the same model with JUST wetlands for alpha S
?lm
canopy_model <- lm(bird_div_S_a ~ canopy_cover + spp_counts + wet_area, data = env_matrix_wet)

summary(canopy_model)

TukeyHSD(canopy_model)

# creating the same model with just wetlands for S PIE alpha

canopy_model_PIE <- lm(bird_div_S_PIE_a ~ mean_dbh + mean_dist + canopy_cover + canopy_dom + groundcover + spp_counts, data = env_matrix_wet)

summary(canopy_model_PIE)


# playing with some plots

ggplot(data = env_matrix_wet, mapping = aes(mean_dist, bird_div_S_a), colour = site_type) + geom_point() + geom_smooth(method = lm)

# plot of wet area ~ diversity, strong correlation

ggplot(data = env_matrix_wet, mapping = aes(wet_area, bird_div_S_a), colour = site_type) + geom_point() + geom_smooth(method = lm)

#plot of canopy cover classes 
ggplot(data = env_matrix_wet, mapping = aes(canopy_cover, bird_div_S_a), color = canopy_cover) + geom_boxplot() + theme_bw()


#plot of groundcover classes
ggplot(data = env_matrix_wet, mapping = aes(groundcover, bird_div_S_a)) + geom_boxplot() + theme_bw()

#plot of canopy dominant trees S alpha
ggplot(data = env_matrix_wet, mapping = aes(canopy_dom, bird_div_S_a)) + geom_boxplot() + scale_x_discrete(guide = guide_axis(n.dodge = 3)) + theme_bw()


#plot of canopy dominant trees S PIE alpha
ggplot(data = env_matrix_wet, mapping = aes(canopy_dom, bird_div_S_PIE_a)) + geom_boxplot() + scale_x_discrete(guide = guide_axis(n.dodge = 3)) + theme_bw()

#creating JUST wet models within site (halidon, and stono separate)
summary(aov(bird_div_S_a ~ canopy_cover + canopy_dom + groundcover, data = env_matrix_halidon))

#plot of canopy cover classes 
ggplot(data = env_matrix_halidon, mapping = aes(canopy_cover, bird_div_S_a), color = canopy_cover) + geom_boxplot() + theme_bw()

#creating an lm model to test our vegetation survey variables against bird div
summary(lm(bird_div_S_a ~ mean_dbh + mean_dist, data = env_matrix_halidon))

summary(aov(bird_div_S_a ~ canopy_cover + canopy_dom + groundcover, data = env_matrix_stono))
```

```{r}
#making the kml file into a shape file
install.packages("sf")
library(sf)
samples <- st_read("ephem_bird_LiDAR_newsites2022.kml")
# need to drop 3rd dimension
samples2 <- st_zm(samples, drop=T, what='ZM')
st_write(samples2, "ephem_bird_LiDAR_poly_newsites2022_shape.shp")
```

Now that I created a community matrix, and actually found the Shannon-Weiner
Biodiversity Index for each wetland, it was easier to manually plug them into a different
data sheet that contained all of my wetland assessment vegetation data. 
So I read that file in, and create some models, test for normal distribution. 

```{r}

# once I compute shannon-weiner index, It is easiest to manually plug those numbers
# in as a column in my wetland attribute Google Sheets table

# quick ANOVA

summary(aov())

# all my wetland attribute data with the Shanon-Weiner Bio Index listed as "shan_sr"
wet_attribute_only <- read.csv("~/raw_data - just_wet_attribute_2.csv")

#histogram of the shan_sr_2
histogram(~ (shan_sr_2), data = wet_attribute_only)

?t.test

# boxplot of wetland v upland again

boxplot(wet_attribute_only$shan_sr_2 ~ wet_attribute_only$type)
t.test(shan_sr_2 ~ type, data = wet_attribute_only)

ggplot() + geom_boxplot(data = wet_attribute_only, aes(x = type, y = shan_sr_2, colour = type)) + theme_bw()

# checking for a normal distribution, it looks good
shapiro.test(wet_attribute_only$shan_2_log)

shan_2_log <- sqrt(wet_attribute_only$shan_sr_2)
?cbind
wet_attribute_only <- cbind(wet_attribute_only,shan_2_log)

# the actual ANOVA table with all 3 vegetation variables 
sr_22 <- (lm(shan_sr_2 ~ canopy_cover, data = wet_attribute_only))

plot(sr_22)

# none of these are statistically signifcant, but I've got 3 plots to visulaize the data, it's apparent that I don't have great sample sizes at lower canopy class and midstory levels 

# boxplot shan div ~ canopy cover
ggplot(data = wet_attribute_only, 
       mapping = aes(x = canopy_cover, y = shan_sr, colour = canopy_cover)) + 
  geom_point() + geom_boxplot() + theme_bw() + 
  xlab("Canopy Cover (%), broken into 4 categories") +
  ylab("Shannon-Weiner Index of Biodiversity at wetland sites")


# boxplot shan div ~ midstory cover
 ggplot(data = wet_attribute_only, 
       mapping = aes(x = midstory_cover, y = shan_sr_1, colour = midstory_cover)) + 
  geom_point() + geom_boxplot() + theme_bw() + 
  xlab("Midstory Cover (%), broken into 4 categories") +
  ylab("Shannon-Weiner Index of Biodiversity at wetland sites")

 # boxplot shan div ~ herbaceous cover
 ggplot(data = wet_attribute_only, 
       mapping = aes(x = herb_layer, y = shan_sr, colour = herb_layer)) + 
  geom_point() + geom_boxplot() + theme_bw() + 
  xlab("Herb Layer, broken into 4 categories") +
  ylab("Shannon-Weiner Index of Biodiversity at wetland sites")

```

```{r}
# Working on the LiDAR data to create similar diversity ~ canopy cover models
library(lidR)
library(sp)

las <- readLAS('./merged_Stono_all_returns_clipped.las', select='xyzcr' )

## coordinate change to utm for Stono
coords_ll <- las@data[ , c('X', 'Y')]
coords_ll <- SpatialPoints(coords_ll, 
                           proj4string =  CRS("+proj=longlat +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,0"))

#plot(coords_ll)
coords_utm <- sp::spTransform(coords_ll,
                              CRSobj = CRS("+proj=utm +zone=17 +datum=WGS84"))
plot(coords_utm)

las@data$X <- coordinates(coords_utm)[ , 1]
las@data$Y <- coordinates(coords_utm)[ , 2]

#checking to see if the coordinates changed to UTM properly
head(las@data)
table(las@data$Classification)

plot(las)

las_5 <- subset(las)

# str(las)

# las <- readLAS('./USGS_LPC_SC_SavannahPeeDee_2019_B19_10493971.laz')

plot(las)

chm <- rasterize_canopy(las, 2, p2r())
col <- random.colors(50)
plot(chm, col = col)

plot(las, color = "ScanAngleRank", bg = "white", axis = TRUE, legend = TRUE)

plot(las, color ="RGB")
plot(las, color = "Intensity", breaks = "quantile", bg = "white")

x <- plot(las, bg = "white", size = 3)
add_dtm3d(x, dtm)

x <- plot(las, bg = "white", size = 3)
add_treetops3d(x, ttops)

gnd <- filter_ground(las)
plot(gnd, size = 3, bg = "white", color = "Classification")


dtm <- rasterize_terrain(las, 1, knnidw())
plot(dtm, col = gray(1:50/50))

nlas <- las - dtm
plot(nlas, size = 4, bg = "white")
hist(filter_ground(nlas)$Z, main = "", xlab = "Elevation")
hist(nlas$Z)


#smaller ws argument means smaller area to compute local maxima
ttops <- locate_trees(las, lmf(ws = 5))
ttops

plot(chm, col = height.colors(50))
plot(sf::st_geometry(ttops), add = TRUE, pch = 3)

x <- plot(las, bg = "white", size = 4)
add_treetops3d(x, ttops)

#larger ws argument = bigger area to compute local maxima
ttops_11m <- locate_trees(las, lmf(ws = 11))
plot(sf::st_geometry(ttops_11m), add = TRUE, pch = 3)


plot(chm, col = height.colors(50))
plot(las)

plot(sf::st_geometry(ttops), pch = 3)

# normalizig height by creating a dtm, seems like maybe we don't have to do this
dtm <- rasterize_terrain(las, 1, knnidw())
plot(dtm, col = gray(1:50/50))

nlas <- las - dtm
plot(nlas, size = 4, bg = "white")

# checking to see if we actually normalized the height
hist(filter_ground(nlas)$Z, breaks = seq(-0.6, 0.6, 0.01), main = "", xlab = "Elevation")

#
#
#
#
#trying to clip and subset based on plot locations
coords <- data.frame(x = rnorm(100, 0, 1), y = rnorm(100, 0,1))
i=1
#for (i in 1:nrow(coords)) {
  circle_clip <- clip_circle(las_segmented_trees,
                             x =  coords$x[i],
                             y = coords$y[i], radius = 25)
  # extract canopy variables
  
  
#}

UP05 <- clip_circle(las_segmented_trees, x =  575908.12, y = 3622341.77, radius = 25)

SP02 <- clip_circle(las_segmented_trees, x =  575952.22, y = 3622456.3, radius = 25)

SP07 <- clip_circle(las_segmented_trees, x =  576855.42, y = 3623070.8, radius = 25)

SP03 <- clip_circle(las_segmented_trees, x =  576310.19, y = 3622208.49, radius = 25)

SP08 <- clip_circle(las_segmented_trees, x =  576084.84, y = 3622170.06, radius = 25)

SP09 <- clip_circle(las_segmented_trees, x =  576435.93, y = 3623036.52, radius = 25)

SP05 <- clip_circle(las_segmented_trees, x =  576765.98, y = 3623368.34, radius = 25)

SP01 <- clip_circle(las_segmented_trees, x =  576625.84, y = 3622952.62, radius = 25)

UP06 <- clip_circle(las_segmented_trees, x =  576727.02, y = 3623075.35, radius = 25)


#mean tree height
cloud_metrics(UP05, func = ~mean(Z))
cloud_metrics(SP02, func = ~mean(Z))
cloud_metrics(SP07, func = ~mean(Z))
cloud_metrics(SP03, func = ~mean(Z))
cloud_metrics(SP08, func = ~mean(Z))
cloud_metrics(SP09, func = ~mean(Z))
cloud_metrics(SP05, func = ~mean(Z))
cloud_metrics(SP01, func = ~mean(Z))

# canopy cover model

# chm <- rasterize_canopy(las, res = 0.5, algorithm = p2r())

# this one creates a subcircle, to reduce the number of empty pixels
# this is on the normalized height model

chm <- rasterize_canopy(las, res = 0.5, 
                        algorithm = p2r(subcircle = 0.15))
col <- height.colors(25)
plot(chm, col = col)

chm_2 <- rasterize_canopy(nlas, 0.5, pitfree(subcircle = 0.2))
plot(las, bg = "white", size = 4)

# trying to get actual math
tree_metrics(las_segmented_trees, func = .stdmetrics)

metrics <- crown_metrics(las_segmented_trees, func = ccm, geom = "convex")
plot(metrics["z_max"], pal = hcl.colors)

cloud_metrics(las, func = ~mean(Z))

metrics <- cloud_metrics(las, func = .stdmetrics_z)
str(head(metrics)) # output is a list

# trying to segment trees
?segment_trees
las_segmented_trees <- segment_trees(nlas, li2012(R = 3, speed_up = 5))

plot(las_segmented_trees, color = "treeID")

#choosing a random tree, and getting multiple measurments
tree110 <- filter_poi(las_segmented_trees, treeID == 110)
plot(tree110, size = 8, bg = "white")

# tree crowns

crowns<- las_segmented_trees()
crowns <- crown_metrics(las_segmented_trees, 
                        func = .stdtreemetrics, geom = "convex")
plot(crowns["convhull_area"], main = "Crown area (convex hull)")

# this was used in the lidR vignette, got an error message
# algo <- dalponte2016(chm, ttops_11m)
# las <- segment_trees(las, algo) # segment point cloud
# plot(las, bg = "white", size = 4, color = "treeID") # visualize trees
# treetop shiny tool
launchApp()


```
*Doing everything for Halidon*

```{r}

las_hh <- readLAS('./HH_merged_all_returns_clipped.las', select='xyzc')

## coordinate change for Halidon
coords_hh <- las_hh@data[ , c('X', 'Y')]
coords_hh <- SpatialPoints(coords_hh, 
                           proj4string =  CRS("+proj=longlat +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,0"))

#plot(coords_hh)
coords_utm_hh <- sp::spTransform(coords_hh,
                              CRSobj = CRS("+proj=utm +zone=17 +datum=WGS84"))
#plot(coords_utm_hh)

las_hh@data$X <- coordinates(coords_utm_hh)[ , 1]
las_hh@data$Y <- coordinates(coords_utm_hh)[ , 2]

plot(las_hh)

chm_hh <- rasterize_canopy(las_hh, 2, p2r())
col <- random.colors(50)
plot(chm_hh, col = col)

plot(las_hh, bg = "white", axis = TRUE, legend = TRUE)

# theres definitley some error points in here

cloud_metrics(nlas_hh, func = ~max(Z))
# 96.41 m is definitley not the highest tree, so we need to filter

#took everything out below 0, and everything over 50m
las_hh_norm <- filter_poi(las_hh, Z >= 0, Z <= 50)

# normailzig height by creating a dtm, seems like maybe we don't have to do this
nlas_hh <- normalize_height(las_hh_norm, knnidw())
plot(nlas_hh, size = 4, bg = "white")

#segmenting trees
las_segmented_trees_hh <- segment_trees(nlas_hh, li2012(R = 3, speed_up = 5))

plot(las_segmented_trees_hh, color = "treeID")

# ##########
# clipping the .las to individual wetlands
HH22 <- clip_circle(las_segmented_trees_hh, x =  610406.39	, y = 3658329.59, radius = 25)

HH17 <- clip_circle(las_segmented_trees_hh, x =  610154.72	, y = 3658456.49, radius = 25)

HH14 <- clip_circle(las_segmented_trees_hh, x =  609835.04	, y = 3658404.12, radius = 25)

HH50 <- clip_circle(las_segmented_trees_hh, x =  609042.44	, y = 3657299.73, radius = 25)

HH48 <- clip_circle(las_segmented_trees_hh, x =  609012.22	, y = 3656910.19, radius = 25)

HH49 <- clip_circle(las_segmented_trees_hh, x =  608983.27	, y = 3656994.14, radius = 25)

HH07 <- clip_circle(las_segmented_trees_hh, x =  609592.11	, y = 3659169.83, radius = 25)

HH02 <- clip_circle(las_segmented_trees_hh, x =  607877.89	, y = 3659680.84, radius = 25)

HH04 <- clip_circle(las_segmented_trees_hh, x =  608471.13	, y = 3659729.52, radius = 25)

HH34 <- clip_circle(las_segmented_trees_hh, x =  609706.05	, y = 3657831.62, radius = 25)

HH13 <- clip_circle(las_segmented_trees_hh, x =  609560.69	, y = 3657970.81, radius = 25)

HH33 <- clip_circle(las_segmented_trees_hh, x =  609759.47	, y = 3657648.15, radius = 25)

# computing the means 
cloud_metrics(HH22, func = ~mean(Z))
cloud_metrics(HH17, func = ~mean(Z))
cloud_metrics(HH14, func = ~mean(Z))
cloud_metrics(HH50, func = ~mean(Z))
cloud_metrics(HH48, func = ~mean(Z))
cloud_metrics(HH49, func = ~mean(Z))
cloud_metrics(HH07, func = ~mean(Z))
cloud_metrics(HH02, func = ~mean(Z))
cloud_metrics(HH04, func = ~mean(Z))
cloud_metrics(HH34, func = ~mean(Z))
cloud_metrics(HH13, func = ~mean(Z))
cloud_metrics(HH33, func = ~mean(Z))


## creating a column of the tree heights 
tree_height_lidar <- c(6.294029, 5.730147, 9.405665, 8.455718,
                       6.337186, 9.038951, 8.733791, NA, NA, 
                       10.37293, 5.043724, 11.58307, 7.829237,
                       7.829237, 5.059275, 6.577072, 5.882933,
                       11.58944, 5.521941, 7.56211, 4.156456)
wet_attribute_only <- cbind(wet_attribute_only, tree_height_lidar)

?cbind

```

```{r}
# creating an lm model with the tree heights as continuous

summary(lm(shan_sr ~ tree_height_lidar, data = wet_attribute_only))

shapiro.test(wet_attribute_only$tree_height_lidar)

ggplot(data = wet_attribute_only, 
       mapping = aes(x = tree_height_lidar, y = shan_sr)) + geom_point() +  geom_smooth(method = lm) + theme_bw() + xlab(" Mean Tree Height (m)") + ylab("Shannon-Weiner Index of Biodiversity at wetland sites")

```

```{r}
# trying to get an rda with env ~ comm
# subset df3 to only get the wetlands
bird_matrix <- df3[1:19, ]

#trying to make the env matrix same size

#got just the wetlands with no NAs
wet_1 <- wet_attribute_only[c(1:7, 10:21), ]
wet_1_updated <- wet_1[order(wet_1$wetland_id), ]

#take out stuff that I don't need
wet_2 <- wet_1_updated[, c(13, 16:19, 22, 26, 44)]
wet_3 <- wet_2[ , 2:8]

wet_3$canopy_cover <- factor(wet_3$canopy_cover)
wet_3$canopy_dom <- factor(wet_3$canopy_dom)
wet_3$midstory_cover <- factor(wet_3$midstory_cover)
wet_3$midstory_dom <- factor(wet_3$midstory_dom)
wet_3$herb_layer <- factor(wet_3$herb_layer)
#actually making the freaking rda and plotting it
rda_birds <- rda(bird_matrix ~ wet_3$tree_height_lidar + wet_3$herb_layer + wet_3$canopy_dom + wet_3$midstory_dom)
rda_birds

plot(rda_birds)

RsquareAdj(rda_birds)

anova(rda_birds, by='margin', permutations=100)

canopy_cover <- dummy(wet_3$canopy_cover)


```

```{r}
plot(rda_birds, type='n', scaling=1)
orditorp(rda_birds, display='sp', cex=0.5, scaling=1, col='blue')
text(rda_birds, display='cn', col='red')
```

