---
title: "bird_quant_analysis"
author: "Jackson Barratt Heitmann"
date: "4/4/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

So this first section I'm taking my raw data, and manipulating it into a community
matrix that I can perform some statistics on afterward. 

```{r}
library(lattice)
library(ggplot2)
library(tidyr)
library(dplyr)
library(readr)
library(mobr)
library(vegan)
library(lidR)
# library(devtools)
# install.packages("devtools")
# install.packages("devtools.zip", repos = NULL)
library(remotes)


#all the bird data together

dat_all <- read.csv("raw_data - all_bird_data.csv")

# env matrix
bird_env <- read.csv("raw_data - veg_2022.csv")

# wetland assessment env matrix

bird_wet_env <- read.csv("raw_data - just_wet_attribute_final.csv")

bird_wet_env <- cbind(bird_wet_env, lidar_tree_metrics)

# renaming the env dataframe for uniola database

super_bird_matrix <- merge(dat_all, bird_env, by = 'wetland_id')

super_bird_matrix_wet <- merge(super_bird_matrix, bird_wet_env, by = 'wetland_id')


# all data together, need to make super matrix, merge()
# with env data, then separate out, individually 
# env matrix, needs site, year, site_type, etc. 

id3 <- with(super_bird_matrix, paste(wetland_id, date, year))
comm3 <- with(super_bird_matrix_wet, tapply(total, list(id3, species), sum))
comm3 <- ifelse(is.na(comm3), 0, comm3)

# create columns for the group variables by splitting row names up
comm3 <- data.frame(comm3)

#I need to fi this in above code, for some reason it summed 4x ??
comm4 <- comm3/4

#to get the raw abundances of all observed species
abun_comm <- comm4 %>%
    summarize_if(is.numeric, sum, na.rm=TRUE)

abun_comm <- as.data.frame(t(abun_comm))
colnames(abun_comm) <- c("Total Abundance")

# getting the percentage of counts that a species was detceted
perc_count <- (colSums(!comm4==0)/nrow(comm4))*100


perc_count <- as.data.frame(perc_count)
colnames(perc_count) <- c("% Counts Detected")

abun_comm <- cbind(abun_comm, perc_count)

# comm3$wetland_id <- sapply(strsplit(row.names(comm3), ' '), function(x) x[1])
# comm3$date <- sapply(strsplit(row.names(comm3), ' '), function(x) x[2])
# comm3$year <- sapply(strsplit(row.names(comm3), ' '), function(x) x[3])

env <- super_bird_matrix_wet %>% select('wetland_id', 'block.y', 'date.x', 'year', 'spp', 'dbh', 'dist', 'site.y', 'site_type.x', 'canopy_cover', 'canopy_dom', 'midstory_cover', 'midstory_dom', 'groundcover', 'dom_herb_group', 'arc_area', 'arc_perim', 'tree_height_mean', 'tree_height_sd', 'crown_hull_mean')


##making the env matrix properly, with spp_richness as a column, with average dbh and dist

env$site_type.x <- as.factor(env$site_type.x)

nrow(distinct(env))
summary(env$site_type.x)

env_species_counts <- env %>% group_by(wetland_id, site.y, year, date.x, block.y, canopy_cover, canopy_dom, midstory_cover, midstory_dom, groundcover, dom_herb_group, arc_area, arc_perim, tree_height_mean, tree_height_sd, crown_hull_mean) %>% summarise(
  spp_counts = length(unique(spp)),
  mean_dbh = mean(dbh, na.rm = T),
  mean_dist = mean(dist, na.rm = T),
  site_type = unique(as.character(site_type.x)))


summary(env$site_type.y)

## making the row names the same
rownames(env_species_counts) <- rownames(comm4)

## do some vegan analysis

rda_tree <- rda(comm4 ~ env_species_counts$spp_counts + env_species_counts$mean_dbh + env_species_counts$canopy_cover + env_species_counts$canopy_dom, data=env_species_counts)
rda_tree

RsquareAdj(rda_tree)

plot(rda_tree, type='n', scaling=1)
orditorp(rda_tree, display='sp', cex=0.5, scaling=1, col='blue')
text(rda_tree, display='cn', col='red')


anova(rda_tree, by='margin', permutations=1000)


# making a mob with env and comm matrix


bird_mob <- make_mob_in(comm4, env_species_counts)


mob_test <- get_delta_stats(bird_mob, env_var = 'site_type', ref_level = 'upland',
                            type = 'discrete', tests = c('SAD', 'N'), n_perm = 199)
                                
plot(mob_test)
# making the alpha, beta, gamme plots, these are now obsolete
# they don't work

bird_stats <- get_mob_stats(bird_mob, group_var = "site_type", ref_level = 'upland', n_perm = 199)


plot(bird_stats, 'S')
plot(bird_stats, 'N')
plot(bird_stats, 'S_n')
plot(bird_stats, 'S_PIE')

# trying to calculate alpha diversity and beta diversity to export
?mobr
bird_div_metrics_S <- calc_comm_div(comm4, 'S', scales = c("alpha"))

plot(bird_div_metrics_S)

bird_div_metrics_S_PIE <- calc_comm_div(comm4, 'S_PIE', scales = c("alpha"))

bird_div_metrics_S_b <- calc_comm_div(comm4, 'S', scales = c("beta"))

bird_div_metrics_S_PIE_b <- calc_comm_div(comm4, 'S_PIE', scales = c("beta"))
 
bird_div_metrics_S_g <- calc_comm_div(comm4, 'S', scales = c("gamma"))

rownames(bird_div_metrics_S) <- rownames(comm4)
rownames(bird_div_metrics_S_PIE) <- rownames(comm4) 
rownames(bird_div_metrics_S_b) <- rownames(comm4)
rownames(bird_div_metrics_S_PIE_b) <- rownames(comm4) 

# try to merge these statistics with my env dataframe, 
# to do some regression modelling

env_matrix <- cbind(env_species_counts, bird_div_metrics_S$value, bird_div_metrics_S_PIE$value)
names(env_matrix)[21] <- 'bird_div_S_a'
names(env_matrix)[22] <- 'bird_div_S_PIE_a'

# trying the gamma diversity extraction by wetland and date


gamma_S <- data.frame(comm4) %>%
  group_by(env_matrix$wetland_id, env_matrix$date.x) %>%
  group_map(~ calc_comm_div(.x, index = c('S'), 
  scales = c("gamma"))) %>%
  bind_rows(.id = 'wetland_id')
gamma_S$wetland_id <- unique(env_matrix$wetland_id)


gamma_df <- data.frame()
# first loop through sites
uni_sites <- unique(env_matrix$wetland_id)
uni_years <- unique(env_matrix$year)
for (i in seq_along(uni_sites)) { 
    #loop through years
    for (j in seq_along(uni_years)) {
        row_indices <- env_matrix$wetland_id == uni_sites[i] &
                       env_matrix$year == uni_years[j]
        if (any(row_indices)) { 
            comm_tmp <- comm4[row_indices, ]
            n_samp <- nrow(comm_tmp )
            N <- sum(comm_tmp)
            alpha_bar <- mean(calc_comm_div(comm_tmp , 'S', scales = 'alpha')$value)
            alpha_S_n <- mean(calc_comm_div(comm_tmp , 'S_n', effort = c(5, 10),
                              scales = 'alpha')$value)
            alpha_PIE <- mean(calc_comm_div(comm_tmp , 'S_PIE', scales = 'alpha')$value, 
                          na.rm = TRUE)
            gamma <- calc_comm_div(comm_tmp , 'S', scales = 'gamma')$value
            tmp_beta_C <- calc_comm_div(comm_tmp , 'S', scales = 'beta')$value[2]
            # need to aggregate env down to a single row and since all values
            # are identical across resurveys I just pull first row of env matrix.
            env_names <- c('canopy_cover', 'canopy_dom', 'arc_area', 
                           'arc_perim', 'spp_counts','mean_dbh',
                           'mean_dist', 'site_type', 'tree_height_mean', 
                           'tree_height_sd','crown_hull_mean')
            env_tmp <- as.data.frame(env_matrix[which(row_indices)[1] , env_names])
            # build output data.frame
            gamma_df <- rbind(gamma_df, data.frame(wetland_id = uni_sites[i],
                                                   year = uni_years[j],
                                                   n_samp, N, gamma, alpha_bar, alpha_S_n,
                                                   alpha_PIE,
                                                   tmp_beta_C, env_tmp))
        } 
    } 
}

head(gamma_df)

?calc_comm_div

#take gamma_df and get just wetlands
gamma_df_wet <- gamma_df[1:50, ]
head(gamma_df_wet)
gamma_df$site_type <- as.factor(gamma_df$site_type)


#lets make a model using the NEWEST env matrix gamma_df_wet, alpha

summary(lm(alpha_bar ~ canopy_cover +
             arc_area + spp_counts + crown_hull_mean, data = gamma_df))

#
summary(aov(alpha_bar ~ site_type, data = gamma_df))

#abundance between site types
boxplot(N ~ site_type, data = gamma_df)
summary(lm(N ~ site_type, data = gamma_df))

#alpha diversity between site types
boxplot(alpha_bar ~ site_type, data = gamma_df)
summary(lm(alpha_bar ~ site_type, data = gamma_df))

#alpha S_n diversity between site types
boxplot(alpha_S_n ~ site_type, data = gamma_df)
summary(lm(alpha_S_n ~ site_type, data = gamma_df))

#alpha S_PIE diversity between site types
boxplot(alpha_PIE ~ site_type, data = gamma_df)
summary(lm(alpha_PIE ~ site_type, data = gamma_df))

#beta diversity between site types
boxplot(tmp_beta_C ~ site_type, data = gamma_df)

#gamma diversity between site types
boxplot(gamma ~ site_type, data = gamma_df)

#lets make a model using the NEWEST env matrix gamma_df_wet, gamma

summary(lm(gamma ~ canopy_cover + mean_dbh +
             arc_area + spp_counts, data = gamma_df_wet))

#plot with arc_area 
ggplot(data = gamma_df_wet, mapping = aes(arc_area, alpha_bar), colour = site_type) + geom_point() + geom_smooth(method = lm)

#plot with spp_counts (aka. tree species richness)
ggplot(data = gamma_df_wet, mapping = aes(spp_counts, alpha_bar), colour = site_type) + geom_point() + geom_smooth(method = lm)

#plot with crown_hull_mean 
ggplot(data = gamma_df_wet, mapping = aes(crown_hull_mean, alpha_bar), colour = site_type) + geom_point() + geom_smooth(method = lm)

#plot with canopy cover class
ggplot(data = gamma_df_wet, mapping = aes(canopy_cover, alpha_bar), colour = canopy_cover) +
  geom_boxplot()

# helpful logical functions: any, all
# helpful character functions: identical, duplicates, 
all(c(F, F, F, F))


gamma_S_PIE <- data.frame(comm4) %>%
  group_by(env_matrix$wetland_id, env_matrix$date.x) %>%
  group_map(~ calc_comm_div(.x, index = c('S_PIE'), 
  scales = c("gamma"))) %>%
  bind_rows(.id = 'wetland_id')

#chaninging rownames to match the community matrix
rownames(gamma_S) <- rownames(comm4)
rownames(gamma_S_PIE) <- rownames(comm4)


```



```{r}
#making the kml file into a shape file
install.packages("sf")
library(sf)
samples <- st_read("ephem_bird_LiDAR_newsites2022.kml")
# need to drop 3rd dimension
samples2 <- st_zm(samples, drop=T, what='ZM')
st_write(samples2, "ephem_bird_LiDAR_poly_newsites2022_shape.shp")
```

Now that I created a community matrix, and actually found the Shannon-Weiner
Biodiversity Index for each wetland, it was easier to manually plug them into a different
data sheet that contained all of my wetland assessment vegetation data. 
So I read that file in, and create some models, test for normal distribution. 

```{r}

#####PRELIM ANALYSIS--NOT SURE I NEED




#binding the values to the env matrix
env_matrix <- cbind(env_matrix, gamma_S$value, gamma_S_PIE$value)
names(env_matrix)[20] <- 'bird_div_gamma_S'
names(env_matrix)[21] <- 'bird_div_gamma_S_PIE'


#checking for normality, bird_div_S_a is good, but S_PIE needs
# a log transformation

hist(env_matrix$bird_div_S_a)

hist(env_matrix_wet$wet_area)

shapiro.test(env_matrix_wet$wet_area)
qqplot(canopy_model$residuals, canopy_model$fitted.values)

# making NAs into 0s
env_matrix[is.na(env_matrix)] <- 0

# need to subset to just get wetlands for some analysis

env_matrix_wet <- subset(env_matrix, site_type == 'wetland')

# sub-setting the JUST wetlands by site, to compare within site variation

env_matrix_halidon <- subset(env_matrix_wet, site.y == 'halidon')

# sub-setting the JUST wetlands by site, to compare within site variation

env_matrix_stono <- subset(env_matrix_wet, site.y == 'stono')


# Model that comparies bird div between site and site type
summary(aov(bird_div_S_a ~ site_type + site.y + canopy_cover, data = env_matrix))


#plot between wetland v upland
ggplot(data = env_matrix, mapping = aes(site_type, bird_div_S_a), color = site_type) + geom_boxplot() + theme_bw()

# plot between stono v halidon
ggplot(data = env_matrix, mapping = aes(site.y, bird_div_S_a), color = site.y) + geom_boxplot() + theme_bw()

# creating the same model with JUST wetlands for alpha S
?lm
canopy_model <- lm(bird_div_S_a ~ canopy_cover + spp_counts + wet_area, data = env_matrix_wet)

summary(canopy_model)

TukeyHSD(canopy_model)

# creating the same model with just wetlands for S PIE alpha

canopy_model_PIE <- lm(bird_div_S_PIE_a ~ mean_dbh + mean_dist + canopy_cover + canopy_dom + groundcover + spp_counts, data = env_matrix_wet)

summary(canopy_model_PIE)

# creating the same model with just wetlands for gamma S

canopy_model_gamma_S <- lm(bird_div_gamma_S ~ canopy_cover + spp_counts + wet_area, data = env_matrix_wet)

summary(canopy_model_gamma_S)

# creating the same model with just wetlands for gamma S_PIE
canopy_model_gamma_S_PIE <- lm(bird_div_gamma_S_PIE ~ canopy_cover + spp_counts + wet_area, data = env_matrix_wet)

summary(canopy_model_gamma_S_PIE)

# playing with some plots

ggplot(data = env_matrix_wet, mapping = aes(mean_dist, bird_div_S_a), colour = site_type) + geom_point() + geom_smooth(method = lm)

# plot of wet area ~ S alpha, strong correlation

ggplot(data = env_matrix_wet, mapping = aes(wet_area, bird_div_S_a), colour = site_type) + geom_point() + geom_smooth(method = lm)

#plot of wet area ~ gamma S

ggplot(data = env_matrix_wet, mapping = aes(wet_area, bird_div_gamma_S), colour = site_type) + geom_point() + geom_smooth(method = lm)

#plot of canopy cover classes 
ggplot(data = env_matrix_wet, mapping = aes(canopy_cover, bird_div_S_a), color = canopy_cover) + geom_boxplot() + theme_bw()


#plot of groundcover classes
ggplot(data = env_matrix_wet, mapping = aes(groundcover, bird_div_S_a)) + geom_boxplot() + theme_bw()

#plot of canopy dominant trees S alpha
ggplot(data = env_matrix_wet, mapping = aes(canopy_dom, bird_div_S_a)) + geom_boxplot() + scale_x_discrete(guide = guide_axis(n.dodge = 3)) + theme_bw()


#plot of canopy dominant trees S PIE alpha
ggplot(data = env_matrix_wet, mapping = aes(canopy_dom, bird_div_S_PIE_a)) + geom_boxplot() + scale_x_discrete(guide = guide_axis(n.dodge = 3)) + theme_bw()

#creating JUST wet models within site (halidon, and stono separate)
summary(aov(bird_div_S_a ~ canopy_cover + canopy_dom + groundcover, data = env_matrix_halidon))

#plot of canopy cover classes 
ggplot(data = env_matrix_halidon, mapping = aes(canopy_cover, bird_div_S_a), color = canopy_cover) + geom_boxplot() + theme_bw()

#creating an lm model to test our vegetation survey variables against bird div
summary(lm(bird_div_S_a ~ mean_dbh + mean_dist, data = env_matrix_halidon))

summary(aov(bird_div_S_a ~ canopy_cover + canopy_dom + groundcover, data = env_matrix_stono))

# once I compute shannon-weiner index, It is easiest to manually plug those numbers
# in as a column in my wetland attribute Google Sheets table

# quick ANOVA

summary(aov())

# all my wetland attribute data with the Shanon-Weiner Bio Index listed as "shan_sr"
wet_attribute_only <- read.csv("~/raw_data - just_wet_attribute_2.csv")

#histogram of the shan_sr_2
histogram(~ (shan_sr_2), data = wet_attribute_only)

?t.test

# boxplot of wetland v upland again

boxplot(wet_attribute_only$shan_sr_2 ~ wet_attribute_only$type)
t.test(shan_sr_2 ~ type, data = wet_attribute_only)

ggplot() + geom_boxplot(data = wet_attribute_only, aes(x = type, y = shan_sr_2, colour = type)) + theme_bw()

# checking for a normal distribution, it looks good
shapiro.test(wet_attribute_only$shan_2_log)

shan_2_log <- sqrt(wet_attribute_only$shan_sr_2)
?cbind
wet_attribute_only <- cbind(wet_attribute_only,shan_2_log)

# the actual ANOVA table with all 3 vegetation variables 
sr_22 <- (lm(shan_sr_2 ~ canopy_cover, data = wet_attribute_only))

plot(sr_22)

# none of these are statistically signifcant, but I've got 3 plots to visulaize the data, it's apparent that I don't have great sample sizes at lower canopy class and midstory levels 

# boxplot shan div ~ canopy cover
ggplot(data = wet_attribute_only, 
       mapping = aes(x = canopy_cover, y = shan_sr, colour = canopy_cover)) + 
  geom_point() + geom_boxplot() + theme_bw() + 
  xlab("Canopy Cover (%), broken into 4 categories") +
  ylab("Shannon-Weiner Index of Biodiversity at wetland sites")


# boxplot shan div ~ midstory cover
 ggplot(data = wet_attribute_only, 
       mapping = aes(x = midstory_cover, y = shan_sr_1, colour = midstory_cover)) + 
  geom_point() + geom_boxplot() + theme_bw() + 
  xlab("Midstory Cover (%), broken into 4 categories") +
  ylab("Shannon-Weiner Index of Biodiversity at wetland sites")

 # boxplot shan div ~ herbaceous cover
 ggplot(data = wet_attribute_only, 
       mapping = aes(x = herb_layer, y = shan_sr, colour = herb_layer)) + 
  geom_point() + geom_boxplot() + theme_bw() + 
  xlab("Herb Layer, broken into 4 categories") +
  ylab("Shannon-Weiner Index of Biodiversity at wetland sites")

```

```{r}
# Working on the LiDAR data to create similar diversity ~ canopy cover models
# install.packages("lidR")
# install.packages("terra")
# install.packages("rgdal")
library(lidR)
library(sp)
library(terra)
library(rgdal)
las <- readLAS("Stono_Preserve_all_returns.las", select='xyzcr' )
wetland_coords <- read.csv("raw_data - wetland_coords.csv")

## coordinate change to utm for Stono
coords_ll <- las@data[ , c('X', 'Y')]
coords_ll <- SpatialPoints(coords_ll, 
                           proj4string =  CRS("+proj=longlat +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,0"))

# plot(coords_ll)
coords_utm <- sp::spTransform(coords_ll,
                              CRSobj = CRS("+proj=utm +zone=17 +datum=WGS84"))
# plot(coords_utm)

las@data$X <- coordinates(coords_utm)[ , 1]
las@data$Y <- coordinates(coords_utm)[ , 2]

#https://epsg.io/32617#:~:text=WGS%2084%20%2F%20UTM%20zone%2017N%20%2D%20EPSG%3A32617
st_crs(las) <- 32617

#checking to see if the coordinates changed to UTM properly
head(las@data)
table(las@data$Classification)

plot(las)

las_5 <- subset(las)

# str(las)

# las <- readLAS('./USGS_LPC_SC_SavannahPeeDee_2019_B19_10493971.laz')


# normalizig height by creating a dtm, 
dtm <- rasterize_terrain(las, 1, knnidw())
# plot(dtm, col = gray(1:50/50))

nlas <- las - dtm
# plot(nlas, size = 4, bg = "white")

# checking to see if we actually normalized the height
hist(filter_ground(nlas)$Z, breaks = seq(-0.6, 0.6, 0.01), main = "", xlab = "Elevation")

#now normalized need to rasterize the canopy

chm_stono <- rasterize_canopy(nlas, 1, p2r(0.15))

ttops <- locate_trees(nlas, lmf(4), uniqueness = "incremental")

algo <- dalponte2016(chm_stono, ttops)
head(algo)
ctg_segmented <- segment_trees(nlas, algo)

# trying to get the crown metrics instead of the cloud metrics
metrics <- crown_metrics(circle_LAS_by_wet[[1]], .stdtreemetrics , geom = "convex")

# plot(metrics["Z"], pal = hcl.colors)

# try to run a for loop that derives canopy variables 
circle_LAS_by_wet <- clip_circle(ctg_segmented,
                                 x = wetland_coords$utm_easting[stono_indices],
                                 y = wetland_coords$utm_northing[stono_indices],
                                 radius = 50)

stono_crown_metrics <- lapply(circle_LAS_by_wet, crown_metrics, .stdtreemetrics , geom = "convex")
stono_tree_height_mean <- unlist(lapply(stono_crown_metrics, function(x) mean(x$Z)))
stono_tree_height_mean <- as.data.frame(stono_tree_height_mean)

stono_tree_height_sd <- unlist(lapply(stono_crown_metrics, function(x) sd(x$Z)))
stono_tree_height_sd <- as.data.frame(stono_tree_height_sd)

stono_crown_hull_area_mean <- unlist(lapply(stono_crown_metrics, function(x) mean(x$convhull_area)))
stono_crown_hull_area_mean <- as.data.frame(stono_crown_hull_area_mean)

stono_tree_metrics <- cbind(stono_tree_height_mean, stono_tree_height_sd, stono_crown_hull_area_mean)
as.data.frame(stono_tree_metrics)
colnames(stono_tree_metrics) <- c("tree_height_mean", "tree_height_sd", "crown_hull_mean")

# try estimating forest cover
n_pts <- 1e4

tst <- as(stono_crown_metrics[[1]], 'Spatial')
box <- bbox(tst)
rnd_pts <- data.frame(x = runif(n_pts, box[1,1], box[1,2]),
                      y = runif(n_pts, box[2,1], box[2,2]))

# need to either only simulate points within the circle or 
# drop those points greater than radius of circle from center
center <- as.data.frame(t(rowMeans(box)))
radius <- 50
pt_dist <- raster::pointDistance(center, rnd_pts, lonlat = FALSE)
rnd_pts <- rnd_pts[pt_dist < radius, ]
# update number of random points
n_pts <- nrow(rnd_pts)
# set projection which is needed for over function
rnd_pts <- SpatialPoints(rnd_pts)
proj4string(rnd_pts) <- crs(tst)
# optional plot to check proof of concept
plot(tst)
points(rnd_pts, pch =19)
text(coordinates(rnd_pts), labels = 1:10, col='red', cex = 1.25)

out <- over(rnd_pts, tst)
out
# forest cover 
(sum(!is.na(out$treeID)) / n_pts) * 100




# lidar_out <- data.frame(wetland_coords[stono_indices, ], mean_tree, sd_tree)
# lidar_out_SP01 <- lidar_out[ 1,]





```
```{r}

##TRYING TO MAKE THIS INTO A FUNCTION

# try estimating forest cover
n_pts <- 1e6

all_canopy_cover <- data.frame()

for (i in seq_along(stono_crown_metrics)){
    tst <- as(stono_crown_metrics[[i]], 'Spatial')
    box <- bbox(tst)
    rnd_pts <- data.frame(x = runif(n_pts, box[1,1], box[1,2]),
                      y = runif(n_pts, box[2,1], box[2,2]))
    
    # need to either only simulate points within the circle or 
    # drop those points greater than radius of circle from center
    center <- as.data.frame(t(rowMeans(box)))
    radius <- 50
    pt_dist <- raster::pointDistance(center, rnd_pts, lonlat = FALSE)
    rnd_pts <- rnd_pts[pt_dist < radius, ]
    # update number of random points
    n_pts <- nrow(rnd_pts)
    # set projection which is needed for over function
    rnd_pts <- SpatialPoints(rnd_pts)
    proj4string(rnd_pts) <- crs(tst)

    out <- over(rnd_pts, tst)
    
    canopy_cover <- data.frame(names(stono_crown_metrics)[i],
                               (sum(!is.na(out$treeID)) / n_pts) * 100)
    
    all_canopy_cover <- rbind(all_canopy_cover, canopy_cover)
    }

colnames(all_canopy_cover) <- c('site', 'canopy_cover')

canopy_cover <- append(canopy_cover, out)
class(out)
```

```{r}
library(lidR)
library(sf)
library(dplyr)

# working example for rasterize_canopy

LASfile <- system.file("extdata", "Topography.laz", package="lidR")
las = readLAS(LASfile, filter = "-inside 273450 5274350 273550 5274450")

col <- pastel.colors(200)

# Points-to-raster algorithm with a resolution of 1 meter
chm <- rasterize_canopy(las, res = 1, p2r())
plot(chm, col = col)

# tree tops
ttops <- locate_trees(chm, lmf(ws = 5))

las_seg <- segment_trees(las, dalponte2016(chm, ttops))
plot(las_seg, color = "treeID")

# calculate crown statistics
crowns <- crown_metrics(las_seg, .stdtreemetrics)

# set reference point as the middle of the bbox
crowns
# 273450.2 ymin: 5274358 xmax: 273549.5 ymax: 5274449

ref_pt <- st_as_sf(data.frame(x = mean(c(273450.2, 273549.5)),
                              y = mean(c(5274358, 5274449))),
                   coords = c('x', 'y'),
                   crs = 2949, remove = FALSE)

crowns_2 <- crowns %>% 
  mutate(distances = st_distance(., ref_pt, by_element = TRUE))

plot(crowns_2)


# convert to spatial points dataframe to make 
# extraction of crown coordinates easier
#crowns_sp <- as(crowns, Class = "Spatial")

hulls <- delineate_crowns(las_seg, func = .stdmetrics)

plot(hulls)


n_pts <- 1e5

bbox(hulls)
rnd_pts <- SpatialPoints(coords = data.frame(x = runif(n_pts, 273450, 273550),
                                             y = runif(n_pts, 5274357, 5274450)))
plot(hulls)
points(rnd_pts, pch =19)
text(coordinates(rnd_pts), labels = 1:10, col='red', cex = 1.25)

proj4string(rnd_pts) <- CRS(proj4string(hulls))

out <- over(rnd_pts, hulls)
out
# forest cover 
(sum(!is.na(out$treeID)) / n_pts) * 100

# need to double check cover estimate here - should be correct but maybe a little too low

# would need to change range of random points to decrease area over which 
# forest cover or other estimates are desired
```



```{r}
# not sure if we need any of this

plot(las)

chm <- rasterize_canopy(las, 2, p2r())
col <- random.colors(50)
plot(chm, col = col)

plot(las, color = "ScanAngleRank", bg = "white", axis = TRUE, legend = TRUE)

plot(las, color ="RGB")
plot(las, color = "Intensity", breaks = "quantile", bg = "white")

x <- plot(las, bg = "white", size = 3)
add_dtm3d(x, dtm)

x <- plot(las, bg = "white", size = 3)
add_treetops3d(x, ttops)

gnd <- filter_ground(las)
plot(gnd, size = 3, bg = "white", color = "Classification")


dtm <- rasterize_terrain(las, 1, knnidw())
plot(dtm, col = gray(1:50/50))

nlas <- las - dtm
plot(nlas, size = 4, bg = "white")
hist(filter_ground(nlas)$Z, main = "", xlab = "Elevation")
hist(nlas$Z)


#smaller ws argument means smaller area to compute local maxima
ttops <- locate_trees(las, lmf(ws = 5))
ttops

plot(chm, col = height.colors(50))
plot(sf::st_geometry(ttops), add = TRUE, pch = 3)

x <- plot(las, bg = "white", size = 4)
add_treetops3d(x, ttops)

#larger ws argument = bigger area to compute local maxima
ttops_11m <- locate_trees(las, lmf(ws = 11))
plot(sf::st_geometry(ttops_11m), add = TRUE, pch = 3)


plot(chm, col = height.colors(50))
plot(las)

plot(sf::st_geometry(ttops), pch = 3)

# then we combine stono and halidon into one data frame

UP05 <- clip_circle(las_segmented_trees, x =  575908.12, y = 3622341.77, radius = 25)

SP02 <- clip_circle(las_segmented_trees, x =  575952.22, y = 3622456.3, radius = 25)

SP07 <- clip_circle(las_segmented_trees, x =  576855.42, y = 3623070.8, radius = 25)

SP03 <- clip_circle(las_segmented_trees, x =  576310.19, y = 3622208.49, radius = 25)

SP08 <- clip_circle(las_segmented_trees, x =  576084.84, y = 3622170.06, radius = 25)

SP09 <- clip_circle(las_segmented_trees, x =  576435.93, y = 3623036.52, radius = 25)

SP05 <- clip_circle(las_segmented_trees, x =  576765.98, y = 3623368.34, radius = 25)

SP01 <- clip_circle(las_segmented_trees, x =  576625.84, y = 3622952.62, radius = 25)

UP06 <- clip_circle(las_segmented_trees, x =  576727.02, y = 3623075.35, radius = 25)


#mean tree height
cloud_metrics(UP05, func = ~mean(Z))
cloud_metrics(SP02, func = ~mean(Z))
cloud_metrics(SP07, func = ~mean(Z))
cloud_metrics(SP03, func = ~mean(Z))
cloud_metrics(SP08, func = ~mean(Z))
cloud_metrics(SP09, func = ~mean(Z))
cloud_metrics(SP05, func = ~mean(Z))
cloud_metrics(SP01, func = ~mean(Z))

# canopy cover model

# chm <- rasterize_canopy(las, res = 0.5, algorithm = p2r())

# this one creates a subcircle, to reduce the number of empty pixels
# this is on the normalized height model

chm <- rasterize_canopy(las, res = 0.5, 
                        algorithm = p2r(subcircle = 0.15))
col <- height.colors(25)
plot(chm, col = col)

chm_2 <- rasterize_canopy(nlas, 0.5, pitfree(subcircle = 0.2))
plot(las, bg = "white", size = 4)

# trying to get actual math
tree_metrics(las_segmented_trees, func = .stdmetrics)

metrics <- crown_metrics(las_segmented_trees, func = ccm, geom = "convex")
plot(metrics["z_max"], pal = hcl.colors)

cloud_metrics(las, func = ~mean(Z))

metrics <- cloud_metrics(las, func = .stdmetrics_z)
str(head(metrics)) # output is a list



#choosing a random tree, and getting multiple measurments
tree110 <- filter_poi(las_segmented_trees, treeID == 110)
plot(tree110, size = 8, bg = "white")

# tree crowns

crowns<- las_segmented_trees()
crowns <- crown_metrics(las_segmented_trees, 
                        func = .stdtreemetrics, geom = "convex")
plot(crowns["convhull_area"], main = "Crown area (convex hull)")

# this was used in the lidR vignette, got an error message
# algo <- dalponte2016(chm, ttops_11m)
# las <- segment_trees(las, algo) # segment point cloud
# plot(las, bg = "white", size = 4, color = "treeID") # visualize trees
# treetop shiny tool
launchApp()

# ##########
# clipping the .las to individual wetlands
HH22 <- clip_circle(las_segmented_trees_hh, x =  610406.39	, y = 3658329.59, radius = 25)

HH17 <- clip_circle(las_segmented_trees_hh, x =  610154.72	, y = 3658456.49, radius = 25)

HH14 <- clip_circle(las_segmented_trees_hh, x =  609835.04	, y = 3658404.12, radius = 25)

HH50 <- clip_circle(las_segmented_trees_hh, x =  609042.44	, y = 3657299.73, radius = 25)

HH48 <- clip_circle(las_segmented_trees_hh, x =  609012.22	, y = 3656910.19, radius = 25)

HH49 <- clip_circle(las_segmented_trees_hh, x =  608983.27	, y = 3656994.14, radius = 25)

HH07 <- clip_circle(las_segmented_trees_hh, x =  609592.11	, y = 3659169.83, radius = 25)

HH02 <- clip_circle(las_segmented_trees_hh, x =  607877.89	, y = 3659680.84, radius = 25)

HH04 <- clip_circle(las_segmented_trees_hh, x =  608471.13	, y = 3659729.52, radius = 25)

HH34 <- clip_circle(las_segmented_trees_hh, x =  609706.05	, y = 3657831.62, radius = 25)

HH13 <- clip_circle(las_segmented_trees_hh, x =  609560.69	, y = 3657970.81, radius = 25)

HH33 <- clip_circle(las_segmented_trees_hh, x =  609759.47	, y = 3657648.15, radius = 25)

# computing the means 
cloud_metrics(HH22, func = ~mean(Z))
cloud_metrics(HH17, func = ~mean(Z))
cloud_metrics(HH14, func = ~mean(Z))
cloud_metrics(HH50, func = ~mean(Z))
cloud_metrics(HH48, func = ~mean(Z))
cloud_metrics(HH49, func = ~mean(Z))
cloud_metrics(HH07, func = ~mean(Z))
cloud_metrics(HH02, func = ~mean(Z))
cloud_metrics(HH04, func = ~mean(Z))
cloud_metrics(HH34, func = ~mean(Z))
cloud_metrics(HH13, func = ~mean(Z))
cloud_metrics(HH33, func = ~mean(Z))


## creating a column of the tree heights 
tree_height_lidar <- c(6.294029, 5.730147, 9.405665, 8.455718,
                       6.337186, 9.038951, 8.733791, NA, NA, 
                       10.37293, 5.043724, 11.58307, 7.829237,
                       7.829237, 5.059275, 6.577072, 5.882933,
                       11.58944, 5.521941, 7.56211, 4.156456)
wet_attribute_only <- cbind(wet_attribute_only, tree_height_lidar)

?cbind


# trying to segment trees at the cloud level
?crown_metrics
las_segmented_trees <- segment_trees(nlas, li2012(R = 3, speed_up = 5))

plot(las_segmented_trees, color = "treeID", colorPalette = col)

#
#
#
#


#re-doing all of this for stono

# so the potential problem with this, is even though we've segmented the trees
# cloud_metrics computes mean and sd for ALL the points
# tree_metrics is the function that actually does this for the segmented trees

stono_indices <- c(21:31, 36:37)
circle_LAS_by_wet <- clip_circle(las_segmented_trees,
                                 x = wetland_coords$utm_easting[stono_indices],
                                 y = wetland_coords$utm_northing[stono_indices],
                                 radius = 50)

names(circle_LAS_by_wet) <- wetland_coords$wetland_id[stono_indices]

mean_tree <- NULL
sd_tree <- NULL

for (i in seq_along(circle_LAS_by_wet)) {
    # extract canopy variables
    mean_tree[i] <- cloud_metrics(circle_LAS_by_wet[[i]], func = ~mean(Z))
    sd_tree[i] <- cloud_metrics(circle_LAS_by_wet[[i]], func = ~sd(Z))
}

lidar_out <- data.frame(wetland_coords[stono_indices, ], mean_tree, sd_tree)
lidar_out_SP01 <- lidar_out[ 1,]


```

*Doing everything for Halidon*

```{r}

las_hh <- readLAS('Hallidon_Hill_all_returns.las', select='xyzc')

## coordinate change for Halidon
coords_hh <- las_hh@data[ , c('X', 'Y')]
coords_hh <- SpatialPoints(coords_hh, 
                           proj4string =  CRS("+proj=longlat +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,0"))

#plot(coords_hh)
coords_utm_hh <- sp::spTransform(coords_hh,
                              CRSobj = CRS("+proj=utm +zone=17 +datum=WGS84"))
#plot(coords_utm_hh)

las_hh@data$X <- coordinates(coords_utm_hh)[ , 1]
las_hh@data$Y <- coordinates(coords_utm_hh)[ , 2]

class(las_hh@data$Y)
plot(las_hh)

sf::sf_use_s2(FALSE)


# normalize height by creating a dtm, seems like maybe we don't have to do this
dtm_hh <- rasterize_terrain(las_hh, 1, knnidw())

?rasterize_terrain

nlas_hh <- las_hh - dtm_hh

chm_hh <- rasterize_canopy(nlas_hh, 1, p2r(0.15))

ttops_hh <- locate_trees(nlas_hh, lmf(4), uniqueness = "incremental")

algo_hh <- dalponte2016(chm_hh, ttops_hh)
head(algo)
ctg_segmented_hh <- segment_trees(nlas_hh, algo_hh)

plot(nlas_hh, size = 4, bg = "white")

# segmenting trees
# las_segmented_trees_hh <- segment_trees(nlas_hh, li2012(R = 3, speed_up = 5))

# plot(las_segmented_trees_hh, color = "treeID")


#trying to clip and subset based on plot locations for halidon


hh_indices <- c(1:20, 32:35)
circle_LAS_by_wet_hh <- clip_circle(ctg_segmented_hh,
                                 x = wetland_coords$utm_easting[hh_indices],
                                 y = wetland_coords$utm_northing[hh_indices],
                                 radius = 25)
names(circle_LAS_by_wet_hh) <- wetland_coords$wetland_id[hh_indices]

# mean_tree_hh <- NULL
# sd_tree_hh <- NULL
# for (i in seq_along(circle_LAS_by_wet_hh)) {
    # extract canopy variables
    # mean_tree_hh[i] <- cloud_metrics(circle_LAS_by_wet_hh[[i]], func = ~mean(Z))
    # sd_tree_hh[i] <- cloud_metrics(circle_LAS_by_wet_hh[[i]], func = ~sd(Z))
#}

# lidar_out_hh <- data.frame(wetland_coords[hh_indices, ], mean_tree_hh, sd_tree_hh)

hh_crown_metrics <- lapply(circle_LAS_by_wet_hh, crown_metrics, .stdtreemetrics , geom = "convex")
hh_tree_height_mean <- unlist(lapply(hh_crown_metrics, function(x) mean(x$Z)))
hh_tree_height_mean <- as.data.frame(hh_tree_height_mean)

hh_tree_height_sd <- unlist(lapply(hh_crown_metrics, function(x) sd(x$Z)))
hh_tree_height_sd <- as.data.frame(hh_tree_height_sd)

hh_crown_hull_area_mean <- unlist(lapply(hh_crown_metrics, function(x) mean(x$convhull_area)))
hh_crown_hull_area_mean <- as.data.frame(hh_crown_hull_area_mean)

hh_tree_metrics <- cbind(hh_tree_height_mean, hh_tree_height_sd, hh_crown_hull_area_mean)

colnames(hh_tree_metrics) <- c("tree_height_mean", "tree_height_sd", "crown_hull_mean")

lidar_tree_metrics <- rbind(stono_tree_metrics, hh_tree_metrics)

```

```{r}
# creating an lm model with the tree heights as continuous

summary(lm(shan_sr ~ tree_height_lidar, data = wet_attribute_only))

shapiro.test(wet_attribute_only$tree_height_lidar)

ggplot(data = wet_attribute_only, 
       mapping = aes(x = tree_height_lidar, y = shan_sr)) + geom_point() +  geom_smooth(method = lm) + theme_bw() + xlab(" Mean Tree Height (m)") + ylab("Shannon-Weiner Index of Biodiversity at wetland sites")

```

```{r}
# trying to get an rda with env ~ comm
# subset df3 to only get the wetlands
bird_matrix <- df3[1:19, ]

#trying to make the env matrix same size

#got just the wetlands with no NAs
wet_1 <- wet_attribute_only[c(1:7, 10:21), ]
wet_1_updated <- wet_1[order(wet_1$wetland_id), ]

#take out stuff that I don't need
wet_2 <- wet_1_updated[, c(13, 16:19, 22, 26, 44)]
wet_3 <- wet_2[ , 2:8]

wet_3$canopy_cover <- factor(wet_3$canopy_cover)
wet_3$canopy_dom <- factor(wet_3$canopy_dom)
wet_3$midstory_cover <- factor(wet_3$midstory_cover)
wet_3$midstory_dom <- factor(wet_3$midstory_dom)
wet_3$herb_layer <- factor(wet_3$herb_layer)
#actually making the freaking rda and plotting it
rda_birds <- rda(bird_matrix ~ wet_3$tree_height_lidar + wet_3$herb_layer + wet_3$canopy_dom + wet_3$midstory_dom)
rda_birds

plot(rda_birds)

RsquareAdj(rda_birds)

anova(rda_birds, by='margin', permutations=100)

canopy_cover <- dummy(wet_3$canopy_cover)


```

```{r}
plot(rda_birds, type='n', scaling=1)
orditorp(rda_birds, display='sp', cex=0.5, scaling=1, col='blue')
text(rda_birds, display='cn', col='red')
```

